Bottum-up drivers of phytoplankton community diversity in the Cariaco Basin, Venezuela

### 2024-12-05 agujournaltemplate.tex
First rough draft of Results

        % below is from 2019 REVIEW
        Hydrographic observations (Figure 2) show strong seasonality and interannual variability in the upper 400 m in the Cariaco Basin. This is controlled by seasonal and interannual changes in the wind and in the intensity of the geostrophic Caribbean Current (Muller-Karger et al. 1989). Primary upwelling was strongest between November and May. In general, during upwelling, sea surface temperatures ranged between approximately 23◦Cand25◦C. The secondary upwelling that occurs between June and August is shorter (∼5 weeks) and is approximately 1.5\°C warmer than the primary upwelling. It is driven by Ekman transport and an intensification of the curl of the wind close to the southern coast of the Caribbean (Rueda-Roa \& Muller-Karger 2013, Rueda-Roa et al. 2018). Upwelled waters move offshore in plumes that extended as far as 50 km to more than 250 km from the coast (Muller-Karger et al. 1989, 2010). From July to October, there is strong stratification in the upper 50 m of the water column after the wind weakens and the rain and rivers deliver less saline waters (Figure 2).
        Sea surface temperatures ranged from approximately 28◦C to 30◦C during the period of stratification (Figure 2). The mixed-layer depth varied from less than 10 m to 35 m during the 21-year time series (Figure 3). Satellite sea surface temperature and ocean color observations show that upwelling pulsates throughout the year at the scale of days to weeks, leading to marked variation in primary production (Figure 3). CARIACO rigorously characterized the annual seasonal variation in primary productivity and stratification. The length of the time series was key in capturing interannual variations that spanned decades. Satellite sea surface temperature and field data show that upwelling was stronger between 1996 and 2002. After 2002, trade wind intensity became weaker (Taylor et al. 2012).
        The field observations, particularly those in the upper 50 m, showed what appeared to be a return to stronger upwelling conditions starting in 2014 (Figure 2a).

@@

% Figure 2
To further investigate the depth distribution and dynamics of chlorophyll biomass and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as across seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters, which during some years is below the concentration between 25 and 50 meters during the years 2000, 2005, 2007, 2008 and 2013 (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a marked reduction in biomass (65\% compared to the mean over the entire time series for these months), which recovers in 2014.

% Figure 3

% Figure 4

% Figure 5

%from Mutshinda et al. 2013
Phytoplankton species richness varied from 4 to 100 species per cruise, with a mean of 40 over the CARIACO time series (167 cruises from November 1995 to March 2011). Over a third of the species (187 of 525) were only observed during one or two cruises in the entire study.

%Text here ===>>>
Below here I will put the draft plots to give an overview over the current state of the results. As of now, all plots are in a draft state, sometimes missing appropriate units. Please excuse this, as I would like to first trim down the plots and order them for the final manuscript, before committing further time into finalizing the details.

>>>>

        % below is from 2019 REVIEW
        Hydrographic observations (Figure 2) show strong seasonality and interannual variability in the upper 400 m in the Cariaco Basin. This is controlled by seasonal and interannual changes in the wind and in the intensity of the geostrophic Caribbean Current (Muller-Karger et al. 1989). Primary upwelling was strongest between November and May. In general, during upwelling, sea surface temperatures ranged between approximately 23\°C and 25\°C. The secondary upwelling that occurs between June and August is shorter ($\sim$ 5 weeks) and is approximately 1.5\°C warmer than the primary upwelling. It is driven by Ekman transport and an intensification of the curl of the wind close to the southern coast of the Caribbean (Rueda-Roa \& Muller-Karger 2013, Rueda-Roa et al. 2018). Upwelled waters move offshore in plumes that extended as far as 50 km to more than 250 km from the coast (Muller-Karger et al. 1989, 2010). From July to October, there is strong stratification in the upper 50 m of the water column after the wind weakens and the rain and rivers deliver less saline waters (Figure 2).
        Sea surface temperatures ranged from approximately 28\°C to 30\°C during the period of stratification (Figure 2). The mixed-layer depth varied from less than 10 m to 35 m during the 21-year time series (Figure 3). Satellite sea surface temperature and ocean color observations show that upwelling pulsates throughout the year at the scale of days to weeks, leading to marked variation in primary production (Figure 3). CARIACO rigorously characterized the annual seasonal variation in primary productivity and stratification. The length of the time series was key in capturing interannual variations that spanned decades. Satellite sea surface temperature and field data show that upwelling was stronger between 1996 and 2002. After 2002, trade wind intensity became weaker (Taylor et al. 2012).
        The field observations, particularly those in the upper 50 m, showed what appeared to be a return to stronger upwelling conditions starting in 2014 (Figure 2a).

@@

% Figure 2
To further investigate the depth distribution and dynamics of chlorophyll biomass and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as across seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters, which during some years is below the concentration between 25 and 50 meters during the years 2000, 2005, 2007, 2008 and 2013 (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% compared to the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. Scaling this genus diversity by the cell count abundances, i.e. the Shannon Index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall. 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The interannual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 3
The effects of diversity on ecosystem functions can be described by the relationship between measured diversity indicators and corresponding measurements of biomass, productivity, and resource use efficiency (RUE). In order to take into account effects of depth on the resulting response, we used the same interpolated depth intervals (0-25, 25-50, 50-75, and 75-100 meters) and plotted coincidental measurements against each other. The considered diversity indices are the genus richness, Shannon index, and Pielou index, which are plotted against chlorophyll a biomass, primary productivity, as well as RUE for Nitrate ($RUE_N$). Immediately apparent, is that depth has a strong effect on the measured relationship, in particular for genus richness, where most instances of high biomass, productivity and $RUE_N$ at high diversity were observed in the uppwer water column (see Figure \ref{fig:bdef} $a, d, g$). For the Shannon index, diversity at the different depths shows a much wider variability and overlap, so the pattern is less clear.  Similarly, community evenness shows no clear depth structure, except that very high biomass and productivity measurements in the upper 25 meters tend to be clustered at intermediate evenness. 
To quantify the relationships, we performed simple linear regression... %TODO perhaps check the glmm method to account for depth and seasonality

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is separated into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
Interannual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlayed with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards postive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. The patterns for upwelling related variables show clear differences between the two clusters, for example for cluster 1 there was a higher occurrence of high wind, shallower 21 \°C Isotherm, as well as lower temperature and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

% Figure 5
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 5 months and MEI v.2 with a lag of 4 months. In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capability for the climate indices with time lags, and upwelling related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower. The 21 \°C Isotherm and silicate and phosphate concentration scored lowest in predicting shifts within the community. The way Gradient Forest predicts shifts in the community, is by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data is biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms. However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the paramater range. The black line indicates the cumulative importance across the entire community, and the colored dashed lines indicate the importance per phytoplankton group. For the AMO index, we see a clear pattern that negative anomalies highly affect the community. This contrasts with a higher importance at positive anomalies in the MEI v.2 index, in particular for Haptophytes. Nitrate concentrations between 6 and 8 $µM$ is where the community changed most, which indicates these values as a threshold for the composition of the community. For wind speed, temperature and salinity, changes occurred most at parameter ranges related to high upwelling, i.e. high wind (between -4 and -6 $m s^{-1}$), low temperature (below 23 \°C) and high salinity (above 36.8 PSU). One interesting aspect is that for both climate indices, Haptophytes showed the strongest cumulative importance, but for Diatoms the nitrate concentration seem to be a stronger predictor.
